---
title: "Housing Status (Test)"
author: "Nikita Voevodin"
date: "3/22/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

installer <- function(x){
  for( i in x ){
    if( !require( i , character.only = TRUE ) ){
      #  If not loading - install
      install.packages( i , dependencies = TRUE )
      #  Load 
      require(i, character.only = TRUE )
    }
  }
}


installer( c("dplyr" , "data.table" , "RMySQL", 'lubridate', 'pbapply') )

```

## Task
The most recent housing status and housing status update date for all clients who were active as of 5-15-2017. It is safe to assume that a client is still active if s/he does not have a subsequent inactive program status update.

## Summary
Here are the results:

* IDS 'JKL', 'STU', and 'PQR' were active as of '2017-05-15' and their latest housing status is 'unhoused';
* ID 'YZA' was active and is housed;
* ID 'VWX' was active but is missing status;
* ID 'GHI' was active but not present at the housing status sheet;
* ID 'STU' is active but is missing program status date (should look into it)

Here is the final table:

```{r echo=FALSE}
print(fread('final_result.csv'))
```

Besides faulty values, there were a couple more things that could be fixed:

* date formats from mdy to ymd;
* column names should all be lower case or at least uniform.


## Analysis

I loaded the data into mySQL test database on AWS so I could show you that I can do both R and SQL, and also for convenience.

Below, you will find 3 different solutions with the same results.

 1. Straight SQL query
 2. R with data pulled from SQL
 3. R with data pulled from CSVs

Note: If you are behind DOITT firewall, the AWS SQL connection might get blocked. You can either switch to a regular WIFI or just go straight to the 3rd solution (R from CSVs).

### Libraries
Installing some of the packages that we might or might not need. I wrote this 'installer' function some time ago to automate library installations. It checks if a library is installed and if yes, it will load it. If not, it will install it and then load it.
```{r eval =FALSE, out.width='80%', fig.asp=.75, fig.align='center'}
installer <- function(x){
  for( i in x ){
    if( !require( i , character.only = TRUE ) ){
      #  If not loading - install
      install.packages( i , dependencies = TRUE )
      #  Load 
      require(i, character.only = TRUE )
    }
  }
}

installer( c("dplyr" , "data.table" , "RMySQL", 'lubridate', 'pbapply') )
```

### SQL Solution

```{r}
# Connecting to the DB:
connection = dbConnect(drv = MySQL(), #specifying database type. 
                       user = "test", # username
                       password = 'testtest', # password
                       host = 'nikitatest.ctxqlb5xlcju.us-east-2.rds.amazonaws.com', # address
                       port = 3306, # port
                       dbname = 'nikita99') # name of the database
```

The query below selects required fields from two derived tables that I joing together using LEFT JOIN. The derived tables are aggregations that return required statues. 

```{r}
resultSQL <- dbGetQuery(connection,
"SELECT 
program.client_id,
program.program_status,
housing.housing_status_date,
housing.housing_status
FROM		
(SELECT 
    table1.client_id, 
    IFNULL(table1.program_status_date, 'missing_date') AS program_status_date,
    table1.program_status 
FROM nikita99.test_program_status as table1 
  INNER JOIN (SELECT 
                  client_id, 
                  MAX(IFNULL(program_status_date, 'missing_date')) AS lastDate
              FROM nikita99.test_program_status
              WHERE program_status_date <= '2017-05-15'
              GROUP BY client_id) AS table2 
    ON table2.client_id = table1.client_id 
    AND table2.lastDate = IFNULL(table1.program_status_date, 'missing_date')
    WHERE program_status IN ('missing_date','active')) AS program   
LEFT JOIN
    (SELECT 
    table3.client_id, 
    table3.housing_status_date,
    IF(table3.housing_status = '', 'missing_status', table3.housing_status) AS housing_status
FROM nikita99.test_housing_status as table3 
  INNER JOIN (SELECT 
                  client_id, 
                  max(housing_status_date) AS lastDate
              FROM nikita99.test_housing_status
             
              GROUP BY client_id) AS table4 
    ON table4.client_id = table3.client_id 
    AND table4.lastDate = table3.housing_status_date)
AS housing
ON housing.client_id = program.client_id
")
# Printing the result
print(resultSQL)
```

### R Solution

```{r message=F}
# Pulling program status table
program <- dbGetQuery(connection,
                      "SELECT *
                      FROM test_program_status")

# Pulling housing status table
housing <- dbGetQuery(connection,
                      "SELECT *
                      FROM test_housing_status")

# Renaming faulty data
program <- setDT(program)[is.na(Program_status_date), Program_status_date:='0000-00-00']

# Using dplyr piping to find the last 
# status by id before or including
# the requested date, and then keep it if it was 'active'.
program <- program %>%
  dplyr::filter(Program_status_date <= '2017-05-15') %>%
  dplyr::group_by(Client_ID) %>% 
  dplyr::arrange(Client_ID, desc(Program_status_date)) %>%
  dplyr::top_n(1) %>%
  dplyr::filter(Program_status == 'Active')

# Keeping the last housing status by id
housing <- housing %>% 
  dplyr::group_by(Client_ID) %>% 
  dplyr::arrange(Client_ID, desc(Housing_Status_Date)) %>%
  dplyr::top_n(1)

# Renaming faulty data
housing <- setDT(housing)[Housing_Status == '', Housing_Status:='missing_status']

# Left joining housing statuses to program statuses to see 
# the housing situation of the active ids. Keeping only relevant
# columns and arranging by date
resultR <- dplyr::left_join(program,housing)[,c(1,2,4,5)] %>% 
  dplyr::arrange(Housing_Status_Date)

# Printing
print(resultR)

# Disconnecting from the DB
dbDisconnect(connection)
```


### R from CSVs
```{r message=F}
# Pulling data from CSVs
housingCSV <- fread('test_housing_status.csv')
programCSV <- fread('test_program_status.csv')

# Dates are messed up. Transforming them to ymd format
programCSV$Program_status_date <- as.character(lubridate::mdy(programCSV$Program_status_date))
housingCSV$Housing_Status_Date <- as.character(lubridate::mdy(housingCSV$Housing_Status_Date))

# Renaming faulty data
programCSV <- setDT(programCSV)[is.na(Program_status_date), Program_status_date:='0000-00-00']

# Using dplyr piping to find the last 
# status by id before or including
# the requested date, and then keep it if it was 'active'.
programCSV <- programCSV %>% 
  dplyr::filter(Program_status_date <= '2017-05-15') %>%
  dplyr::group_by(Client_ID) %>% 
  dplyr::arrange(Client_ID, desc(Program_status_date)) %>%
  dplyr::top_n(1) %>%
  dplyr::filter(Program_status == 'Active')

# Keeping the last housing status by id before or including
# the requested date.
housingCSV <- housingCSV %>% 
  dplyr::group_by(Client_ID) %>% 
  dplyr::arrange(Client_ID, desc(Housing_Status_Date)) %>%
  dplyr::top_n(1)

# renaming empty strings into readable format
housingCSV <- setDT(housingCSV)[Housing_Status == '', Housing_Status:='missing_status']

# Left joining housing statuses to program statuses to see 
# the housing situation of the active ids. Keeping only relevant
# columns and arranging by date
resultR_CSV <- dplyr::left_join(programCSV, housingCSV)[,c(1,2,4,5)] %>% 
  dplyr::arrange(Housing_Status_Date)

# Printing
print(resultR_CSV)

fwrite(resultR_CSV, 'final_result.csv')
```

# THE END